WSClean: an implementation of a fast, generic wide-field imager for radio astronomy

### 2014-07-03 articles/2014/Article-WSClean/wsclean.tex
Fixing suggested edits by reviewer

\date{Accepted TODO. Received TODO; in original form 01-04-2014}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\pubyear{2014}

@@

\subsection{Time complexity of $w$-stacking} \label{sec:time-complexity}
\begin{table}
 \caption{Scaling of the computational cost for various imaging steps, with $N_\textrm{wlay}$ the number of $w$-layers, $N_\textrm{pix}$ the number of pixels along each side, $N_\textrm{vis}$ the number of visibilities, $N_\textrm{kern}$ the size of the anti-aliasing kernel, $N_\textrm{wkern}$ the size of the $w$-kernel, $w_{\max}$ the maximum $w$ value and $\alpha_\textrm{FOV}$ the imaging FOV.} \label{tbl:computational-cost-per-operation}
 \begin{tabular}{lll}
   \textbf{Operation} & \textbf{$w$-stacking} & \textbf{$w$-projection} \\
   \hline\hline
   Fourier transform(s) & $N_\textrm{wlay} N^2_\textrm{pix} \log N_\textrm{pix}$ & $N^2_\textrm{pix} \log N_\textrm{pix}$ \\
   $w$-term corrections   & $N_\textrm{wlay} N^2_\textrm{pix}$ & $N_\textrm{vis} N^2_\textrm{wkern}$ \\
   Gridding & $N_\textrm{vis}N^2_\textrm{kern}$ & $N_\textrm{vis} N_\textrm{kern}^2$ \\
   \hline\hline
 \end{tabular}
\end{table}

@@

As can be seen in Panel~\ref{fig:timing-fov}, the FOV has a large affect on performance when imaging off-zenith. This plot was created by varying the size of a pixel in the output image, such that the number of pixels in the image did not change. The FOV is calculated as the angle subtended between the left- and right-most pixel in the image. \textsc{casa} is not able to make off-zenith images larger than $\sim$30\degree, and its imaging cost increases much more rapidly compared to the cost of \textsc{wsclean}, which implies that gridding is the major cost in this scenario. The image recentring technique is clearly beneficial for FOVs larger than $\sim$30\degree, and can even make a difference of an order of magnitude at very large FOVs of $\sim$90\degree.

@@

The function $t_\textrm{snapshot}$ can be minimised to find the optimal snapshot duration. For \textsc{wsclean} with MWA parameters, we find that this function decreases but no minimum is reached for $\Delta \tau_\textrm{snapshot}<24$~h. Therefore, from a performance perspective, the snapshot duration should be as large as possible. In practise, the beam needs to be corrected on small time scales. A snapshot duration of more than a few minutes is therefore not possible. For $w$-projection in \textsc{casa} we find an optimal snapshot duration of $\sim2$ min for MWA observations.

@@

\subsection{Estimated computing cost for other telescopes} \label{sec:application-to-nonmwa}
We use the derived functions to estimate the computational cost of the algorithms for configurations of several telescopes. We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multi-frequency Snapshot Sky Survey (MSSS)\footnote{See~\href{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}{www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; all Dutch LOFAR stations \citep{lofar-2013}; the Amsterdam--ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \href{http://www.aartfaac.org}{www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); The Meer Karoo Array Telescope (MeerKAT); and the low-frequency Phase~1 aperture arrays of the Square-Kilometre Array (SKA), with the full core (3~km) and the core+arms configurations \citep{ska-phase1-2013}. The configurations are summarised in Table~\ref{tbl:computational-cost-configurations}. In multi-beam configurations such as the EMU and LOFAR configurations, we assume each beam is imaged separately, i.e., the FOV of a single beam is used and the computing cost is multiplied with the number of beams. The selected wavelengths 
are approximately the central wavelength available for each instrument. The image size is set to two times the FOV width divided by the resolution, as described by Eq.~\ref{eq:nyquist-resolution}. Therefore, this assumes that the optimisation of \S\ref{sec:decreasing-inversion-resolution} to inverse at lower resolution is used. The total required computing power is calculated by multiplying the estimated computing time on our test machine with the performance of the machine (138 GFLOPS).

@@

The available SKA computing power is estimated to be around $\sim$100 PetaFLOPS. Extrapolation of our results shows that the current imagers require $3$--$60$ petaFLOPS for SKA1 low alone. Although we have measured our performance in FLOPS, it is likely that memory bandwidth will be a limiting factor, because memory bandwidth is increasing less quickly compared to the floating point performance (e.g., \citealt{gpu-gridding-romein-2012}). \citet{widefield-imaging-ska-cornwell} suggests that the $w$-snapshots algorithm improves the imaging speed for the SKA situation, but our results show that $w$-snapshot imaging does not improve the SKA imaging performance when the $w$-stacking algorithm is used. Clearly it remains challenging to perform wide-field imaging with acceptable performance, but some optimisations could be made for the SKA. For example, a performance gain of factors of a few can be achieved by averaging shorter baselines to their lowest time and bandwidth resolution before imaging. Moreover, although longer baselines make the imaging more expensive, imaging the full FOV will likely not (always) be required when using the longest baselines.

>>>>

\date{Accepted TODO. Received TODO; in original form 1 April 2014}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\pubyear{2014}

@@

\subsection{\editmark{Computational} complexity of $w$-stacking} \label{sec:time-complexity}
\begin{table}
 \caption{Scaling of the computational cost for various imaging steps, with $N_\textrm{wlay}$ the number of $w$-layers, $N_\textrm{pix}$ the number of pixels along each side, $N_\textrm{vis}$ the number of visibilities, $N_\textrm{kern}$ the size of the anti-aliasing kernel, $N_\textrm{wkern}$ the size of the $w$-kernel, $w_{\max}$ the maximum $w$ value and $\alpha_\textrm{FOV}$ the imaging FOV.} \label{tbl:computational-cost-per-operation}
 \begin{tabular}{lll}
   \textbf{Operation} & \textbf{$w$-stacking} & \textbf{$w$-projection} \\
   \hline\hline
   Fourier transform(s) & $N_\textrm{wlay} N^2_\textrm{pix} \log N_\textrm{pix}$ & $N^2_\textrm{pix} \log N_\textrm{pix}$ \\
   $w$-term corrections   & $N_\textrm{wlay} N^2_\textrm{pix}$ & $N_\textrm{vis} N^2_\textrm{wkern}$ \\
   Gridding & $N_\textrm{vis}N^2_\textrm{kern}$ & $N_\textrm{vis} N_\textrm{kern}^2$ \\
   \hline\hline
 \end{tabular}
\end{table}

@@

As can be seen in Panel~\ref{fig:timing-fov}, the FOV has a large \editmark{effect} on performance when imaging off-zenith. This plot was created by varying the size of a pixel in the output image, such that the number of pixels in the image did not change. The FOV is calculated as the angle subtended between the left- and right-most pixel in the image. \textsc{casa} is not able to make off-zenith images larger than $\sim$30\degree, and its imaging cost increases much more rapidly compared to the cost of \textsc{wsclean}, which implies that gridding is the major cost in this scenario. The image recentring technique is clearly beneficial for FOVs larger than $\sim$30\degree, and can even make a difference of an order of magnitude at very large FOVs of $\sim$90\degree.

@@

The function $t_\textrm{snapshot}$ can be minimised to find the optimal snapshot duration. For \textsc{wsclean} with MWA parameters, we find that this function decreases but no minimum is reached for $\Delta \tau_\textrm{snapshot}<24$~h. Therefore, from a performance perspective, the snapshot duration should be as large as possible. In \editmark{practice}, the beam needs to be corrected on small time scales. A snapshot duration of more than a few minutes is therefore not possible. For $w$-projection in \textsc{casa} we find an optimal snapshot duration of $\sim2$ min for MWA observations.

@@

\subsection{Estimated computing cost for other telescopes} \label{sec:application-to-nonmwa}
We use the derived functions to estimate the computational cost of the algorithms for configurations of several telescopes. We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multi-frequency Snapshot Sky Survey (MSSS)\footnote{See~\href{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}{www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; all Dutch LOFAR stations \citep{lofar-2013}; the Amsterdam--ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \href{http://www.aartfaac.org}{www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); The \editmark{MeerKAT}; and the low-frequency Phase~1 aperture arrays of the Square-Kilometre Array (SKA), with the full core (3~km) and the core+arms configurations \citep{ska-phase1-2013}. The configurations are summarised in Table~\ref{tbl:computational-cost-configurations}. In multi-beam configurations such as the EMU and LOFAR configurations, we assume each beam is imaged separately, i.e., the FOV of a single beam is used and the computing cost is multiplied with the number of beams. The selected wavelengths 
are approximately the central wavelength available for each instrument. The image size is set to two times the FOV width divided by the resolution, as described by Eq.~\ref{eq:nyquist-resolution}. Therefore, this assumes that the optimisation of \S\ref{sec:decreasing-inversion-resolution} to \editmark{compute the} inverse at lower resolution is used. The total required computing power is calculated by multiplying the estimated computing time on our test machine with the performance of the machine (138 GFLOPS).

@@

The available SKA computing power is estimated to be around $\sim$100 PetaFLOPS. Extrapolation of our results shows that the current imagers require $3$--$60$ petaFLOPS for SKA1 low alone. Although we have measured our performance in FLOPS, it is likely that memory bandwidth will be a limiting factor, because memory bandwidth is increasing less quickly compared to the floating point performance (e.g., \citealt{gpu-gridding-romein-2012}). \citet{widefield-imaging-ska-cornwell} suggests that the $w$-snapshots algorithm improves the imaging speed for the SKA situation, but our results show that \editmark{$w$-snapshot imaging does not improve SKA imaging performance over $w$-stacking}. Clearly it remains challenging to perform wide-field imaging with acceptable performance, but some optimisations could be made for the SKA. For example, a performance gain of factors of a few can be achieved by averaging shorter baselines to their lowest time and bandwidth resolution before imaging. Moreover, although longer baselines make the imaging more expensive, imaging the full FOV will likely not (always) be required when using the longest baselines.
